Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-02 index.qmd
section:background,dag; process:write

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

@@

<!-- 3. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

@@

<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. @fig-dags illustrates these blocks. @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$. Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$. @fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$. Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely *mediated* by $X_{2}$. @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.

@@

![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=85%}

![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=85%}

![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=85%}

@@

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure. This example can also help to clarify additional conventions for drawing DAGs.

@@

<!-- 5. DAG for the motivating example -->
@fig-example presents two graphs illustrating the plausible causal structure of the motivating example. @fig-dag_example1 shows the *simplified* DAG. In this graph, the variables of interest or *endogenous* variables $V=\{T,X,Y\}$ are depicted as solid black circles, indicating that they are observed variables. The arrows in the graph reflect the expected direction of causal influences among the variables. @fig-dag_example2, in contrast, depicts the *magnified* version of the DAG. This graph includes, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{T},e_{X},e_{Y}\}$. These exogenous variables are commonly referred to as *disturbances* or *errors*, because they represent factors not explicitly modeled. They are depicted as open circles, indicating their unobserved nature. While such exogenous variables are often omitted for simplicity, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts that will be discussed in the next section.

@@

Furthermore, a detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks. @fig-dag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. It also illustrates a fork with $X \rightarrow \{T,Y\}$. Similarly, @fig-dag_example2 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$. The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables. This concept will be introduced in the next section.

@@

@@

>>>>

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

@@

<!-- 3. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

@@

<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020; @Morgan_et_al_2014]. @fig-dags illustrates these blocks. @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$. Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$. @fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$. Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely *mediated* by $X_{2}$. @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.

@@

![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=75%}

![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=75%}

![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=75%}

@@

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure. This example can also help to clarify some additional conventions for drawing DAGs.

@@

<!-- 5. DAG for the motivating example -->
@fig-example presents two graphs illustrating the plausible causal structure of the motivating example. @fig-dag_example1 shows the *simplified* DAG. In this graph, the variables of interest or *endogenous* variables $V=\{T,X,Y\}$ are depicted as solid black circles, indicating that they are observed variables. The arrows in the graph reflect the expected direction of causal influences among the variables. In contrast, @fig-dag_example2 depicts the *magnified* version of the DAG. This graph includes, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{T},e_{X},e_{Y}\}$. These exogenous variables are commonly referred to as *disturbances* or *errors*, because they represent factors not explicitly modeled. They are depicted as open circles, indicating their unobserved nature. These exogenous variables are often omitted for simplicity, resulting in the simplified DAG in @fig-dag_example1. However, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020]. These concepts are introduced in the next section.

@@

Furthermore, a detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks. @fig-dag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. It also illustrates a fork with $X \rightarrow \{T,Y\}$. Similarly, @fig-dag_example2 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$. The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables. This concept is introduced in the next section.

@@

<!-- Assuming certain values are decided in the experimental design (proportions, sample size, etc) you can reduce the graph -->

@@

<!-- Final models does not require the estimation of latent variables, the results can be evaluated without them -->

<!-- Notice that latent variables are kind of like descendants -->
